---
title: "Analyze the diauxic shift expression data using K-means"
author: "Bin He"
date: "2023-06-22 (updated `r Sys.Date()`)"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r load_library, echo=FALSE}
# install the package if not already available
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(ggtext))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(require(broom))
```

## Goal
Learn to use K-means clustering on the diauxic shift expression time course.

## Data
The data are from DeRisi et al. 1997 (DOI: 10.1126/science.278.5338.680). Briefly, researchers sampled yeast grown in the YPD media at 2 hour intervals during a regular growth cycle, labeled the cDNA with Cy3. They used a common reference - the cDNA from the 0 time point, labeled with Cy5. By mixing the latter time point samples each with a copy of the 0 min sample, they are able to get the ratio of expression for each gene, which is presented in log2 transformed forms here.

_Data source_

| GEO# | Description | Reference |
| ---- | ----------- | --------- |
| GSE28 | Microarray expression patterns of _S. cerevisiae_ wt during the metabolic shift from anaerobic fermentation to aerobic respiration. Samples taken at several time points up to 20.5 h. | DeRisi et al. 1997 |

The analysis below is inspired by a GEOquery workshop offered by Jason Ratcliff at the Iowa Institute of Human Genetics. `getGEO()` will return an ExpressionSet object, which is then converted into a "SummarizedExperiment" object, which is a more modern data structure that is easier to deal with.
```{r}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072*10) # increase the local cache size
gse <- GEOquery::getGEO(GEO = "GSE28", destdir = "../input/")
gse <- as(object = gse[[1]], "SummarizedExperiment")
```

_Alternative source_

Instead of processing the GSE data above, I downloaded an expression matrix from this [link](https://www.yeastgenome.org/reference/S000045635) and lightly processed it to remove unnecessary columns and rows, and renamed the column headers
```{r}
raw <- read_tsv("../input/20230622-derisi-1997-expression-matrix.tsv")
```

We will filter this dataset by excluding rows where all time point measurements are within -$\delta$ to $\delta$, with duplicate genes removed
```{r}
delta = 2.2
mat0 = as.matrix(select(raw, -ORF))
not.use = apply(mat0, 1, function(x) all(x > -delta & x < delta) || any(is.na(x)))
use = which(!not.use)
mat = mat0[use,]
rownames(mat) = raw$ORF[use]
all(!duplicated(rownames(mat)))
sprintf("We have a total of %d genes after filtering", nrow(mat))
```
There are no duplicated genes in the filtered set.

## Exploratory Analysis
Let's visualize the data
```{r}
dat.long <- raw[use,] %>% 
  pivot_longer(cols = -ORF, names_to = "time", values_to = "log2") %>% 
  mutate(time = gsub("t", "", time) %>% as.numeric())
```
```{r}
ggplot(dat.long, aes(x = time, y = log2)) + 
  geom_line(aes(group = ORF), linewidth = 0.5, alpha = 0.2)
```
## Clustering
We will use the `kmeans()` function in the `stats` package.
```{r}
# set random seed
set.seed(123)
res <- kmeans(mat, centers = 6, iter.max = 100)
```

Let's look at one of the clusters
```{r}
cluster.labels = tibble(ORF = names(res$cluster), cluster = res$cluster)
left_join(dat.long, cluster.labels, by = "ORF") %>%
  filter(!is.na(cluster)) %>% 
  ggplot(aes(x = time, y = log2)) + 
  geom_line(aes(group = ORF), linewidth = 0.5, alpha = 0.2) +
  facet_wrap(~cluster) + 
  theme_cowplot()
```

